Supplementary Materials

### 2023-06-26 manuscript/supplementary.tex
ms updates

\subsection{YRI Samples}

@@

\subsection{Site-based Filtering of Counts}

@@

\begin{enumerate}
  \item \textbf{Non-accessible regions}: masks out centromeres (\texttt{acen} entries in
    \texttt{cytoBand.txt}), with 5Mbp padding on either side. The file of
    passing masks is \path{data/annotation/no_centro.bed}.

  \item \textbf{Reference masking}: soft and hard-masked regions in the human
      GRCh38 reference genome are also masked. Soft-masked regions were
      determined by Ensembl (XXX), which uses Repeat Masker (XXX) and Dust
      (XXX).

  \item \textbf{Non-``putatively" neutral regions}: Additionally, for fitting our
    likelihood and estimating observed pairwise diversity, we only consider .
    This masks out phastCons regions (from
    \path{phastConsElements100way.bed.gz}) and Ensembl gene regions (from
    annotation file \path{Homo_sapiens.GRCh38.107.chr.gff3.gz}).  While introns
    are possibly under some weak selection, they collectively make up nearly
    40\% of the human genome and are included so genome-wide diversity can be
    estimated more precisely (possibly at the expense of some bias).
\end{enumerate}

These files are all produced by the Snakemake file \path{data/annotation/Snakefile}.
Note that the levels of accessible putatively neutral bases

\begin{figure}[!htb]
  \centering
  \includegraphics{figures/supplementary/chr1_diversity_filtering.pdf}

  \caption{Estimates of chromosome 1 YRI diversity in non-overlapping megabase
    windows, under different filtering criteria. The filtering criteria are,
    (1) ``neutral, accessible" which includes only putatively neutral sites, and
    ignores regions masked as inaccessible, (2) ``accessible" which is only ignores
  sites masked as inaccessible, and (3) ``no filtering" which uses all available
data. Note that filtering changes the absolute level of diversity, but has more
minor effects on regional patterns of diversity at the megabase scale.}

  \label{suppfig:chr1-diversity}
\end{figure}

\begin{table}
\centering
\begin{tabular}{lrrr}
  \hline
chrom &  accessible &  neutral &  both \\
  \hline
 chr1 &        42.4 &     64.4 &  22.9 \\
 chr2 &        46.8 &     58.7 &  23.7 \\
 chr3 &        44.1 &     62.9 &  24.8 \\
 chr4 &        44.0 &     59.3 &  21.5 \\
 chr5 &        44.1 &     58.0 &  21.4 \\
 chr6 &        45.0 &     58.0 &  22.1 \\
 chr7 &        44.4 &     65.8 &  26.4 \\
 chr8 &        43.8 &     61.2 &  23.1 \\
 chr9 &        39.2 &     64.5 &  20.3 \\
chr10 &        44.6 &     62.8 &  25.3 \\
chr11 &        42.7 &     63.5 &  23.4 \\
chr12 &        41.7 &     62.7 &  22.7 \\
chr13 &        39.7 &     62.0 &  18.3 \\
chr14 &        37.6 &     65.4 &  19.0 \\
chr15 &        37.0 &     69.4 &  20.3 \\
chr16 &        39.4 &     63.8 &  20.9 \\
chr17 &        40.4 &     64.0 &  22.3 \\
chr18 &        40.9 &     59.8 &  19.8 \\
chr19 &        30.9 &     69.8 &  18.0 \\
chr20 &        36.9 &     61.6 &  19.8 \\
chr21 &        31.6 &     68.6 &  18.2 \\
chr22 &        26.9 &     75.0 &  16.4 \\
\end{tabular}
\end{table}

>>>>

\subsection{}

@@

\subsection{Non-accessible and Masked Regions}
\label{supp:accessible}

@@

\begin{enumerate}

  \item \textbf{Non-accessible regions}: masks out centromeres (\texttt{acen} entries in
    \texttt{cytoBand.txt}), with 5Mbp padding on either side. The file of
    passing masks is \path{data/annotation/no_centro.bed}.

  \item \textbf{Reference masking}: soft and hard-masked regions in the human
      GRCh38 reference genome are also masked. Soft-masked regions were
      determined by Ensembl (XXX), which uses Repeat Masker (XXX) and Dust
      (XXX). Additionally, we used the 1000 Genomes strict filter (XXX).

\subsection{Putatively Neutral Tracks}

\begin{figure}[!htb]
  \centering
  \includegraphics[width=\textwidth]{figures/supplementary/neutral_tracks.pdf}

  \caption{Estimates of chromosome 1 YRI diversity in non-overlapping megabase
      windows, under different neutral tracks (colors) and the filtering
      criteria described in Supplementary Section \ref{supp:accessible}. To the
      right of the chromosome subfigure are the genome-wide boxplots of
      megabase-scale diversity. }

  \label{suppfig:neutral-tracks}
\end{figure}

Linked selection theory typically assumes that focal sites are not under
purifying selection themselves. Consequently, linked selection statistical
methods often only consider the fraction of the genome not \emph{a priori}
under selective constraint \parencite[e.g. Appendix Section
3.1]{Murphy2022-sj}. Our methods follow suit, and first create a track
consisting of the union (i.e. \texttt{bedtools merge}) of: exons and coding
sequences from Ensembl Version 107
\parencite[\path{Homo_sapiens.GRCh38.107.chr.gff3.gz}]{Cunningham2022-vk} and
PhastCons regions
\parencite[\path{phastConsElements100way.bed.gz}]{Siepel2005-wh}. Then, we
remove 200 basepairs of ``slop" around each of these regions (i.e.
\texttt{bedtools slop}). We show in Supplementary Figure
\ref{suppfig:neutral-tracks} that average diversity (we only consider YRI
samples for this analysis) is fairly insensitive to the slop width, but
increasing slop widths around putatively conserved segments does lead to higher
diversity estimates. This is expected, since increasing the slop width
considers a smaller subset of sites, further away from putatively conserved
regions. For our analyses we choose the 200bp slop version, which covers
63.66\% of the autosomal reference sequence.

Additionally, we compared our neutral track (and the slop width alternatives)
against the bottom 70\% of CADD site scores (this works out to be 64.70\% of
the autosomal reference sequence). The correlation between these two tracks is
extraordinarily high (Pearson $r = 0.98$, p-value $=0$), but the average
pairwise diversity is 13\% higher (two-tailed t-test p-value $1.29 \times
10^{-71}$) in the CADD 70\% neutral track. However, we did not use the CADD
70\%  neutral track in our analysis of human population genomic data. Our
concern was that this neutral track would be too close to the complement of the
CADD 6\% track, which we use as one of our annotation models.

These files are all produced by the Snakemake file
\path{data/fit_annotation/Snakefile} in this study's GitHub repository
(\url{http://github.com/vsbuffalo/bprime}).

\begin{table}
\centering
\caption{XXX}
\begin{tabular}{lrrr}
  \hline
chrom &  accessible &  neutral &  both \\
  \hline
 chr1 &        42.4 &     64.4 &  22.9 \\
 chr2 &        46.8 &     58.7 &  23.7 \\
 chr3 &        44.1 &     62.9 &  24.8 \\
 chr4 &        44.0 &     59.3 &  21.5 \\
 chr5 &        44.1 &     58.0 &  21.4 \\
 chr6 &        45.0 &     58.0 &  22.1 \\
 chr7 &        44.4 &     65.8 &  26.4 \\
 chr8 &        43.8 &     61.2 &  23.1 \\
 chr9 &        39.2 &     64.5 &  20.3 \\
chr10 &        44.6 &     62.8 &  25.3 \\
chr11 &        42.7 &     63.5 &  23.4 \\
chr12 &        41.7 &     62.7 &  22.7 \\
chr13 &        39.7 &     62.0 &  18.3 \\
chr14 &        37.6 &     65.4 &  19.0 \\
chr15 &        37.0 &     69.4 &  20.3 \\
chr16 &        39.4 &     63.8 &  20.9 \\
chr17 &        40.4 &     64.0 &  22.3 \\
chr18 &        40.9 &     59.8 &  19.8 \\
chr19 &        30.9 &     69.8 &  18.0 \\
chr20 &        36.9 &     61.6 &  19.8 \\
chr21 &        31.6 &     68.6 &  18.2 \\
chr22 &        26.9 &     75.0 &  16.4 \\
\end{tabular}
\end{table}
